---
title: "Risk Analysis of Australia's Higher Education Loan Program (HELP)"
author: "Alana Dabelstein, Anh-Thu Hoang, Chenxi Zhang"
date: "01/06/2023"
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    theme: paper
    fig_width: 9
    fig_height: 5
bibliography: references.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Loading Libraries
library(bookdown)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(tidyverse)
library(knitr)
library(gridExtra)
library(eeptools)

# Loading Data
table_1 <- read_xlsx("Data/help-statistics-2021-2022.xlsx", sheet = 2, skip = 2, col_names = TRUE)
table_2 <- read_xlsx("Data/help-statistics-2021-2022.xlsx", sheet = 3, skip = 2, col_names = TRUE)
table_3 <- read_xlsx("Data/help-statistics-2021-2022.xlsx", sheet = 4, skip = 2, col_names = TRUE)
table_4 <- read_xlsx("Data/help-statistics-2021-2022.xlsx", sheet = 5, skip = 2, col_names = TRUE)
table_7 <- read_xlsx("Data/help-statistics-2021-2022.xlsx", sheet = 8, skip = 2, col_names = TRUE)
```

# Introduction

Educational attainment is associated with higher labour force participation, higher income, greater wealth, lower financial stress and healthier behaviours [@oecd_benefits_2018]. Despite this, many individuals have debt-averse attitudes which deter them from applying to university [@c_callender_g_mason_does_2017].

Tertiary education cost has grown fast over time while there has been flat growth in the wages of young adults aged 25–34 [@preston_gender_2023]. The Higher Education Loan Program (HELP) is an income-contingent loan provided by the Australian Government to eligible students to cover this cost.

The income-contingency reduces the risk of borrowing for students where the future value of education or training is uncertain, especially for students from low-income backgrounds [@albrecht1991deferred] as repayment is only required once their income reaches the minimum threshold through the income tax system. Enrolments and the number of people with outstanding HELP debts has steadily increased [@universities_australia_2022_2022]. 

As of 2021-22, the minimum repayment threshold is $47,014. It is projected that HELP debt will total almost $200 billion in 2024–25 and almost one-third will be written off. [@anao2016]. For debtors who do not earn the minimum threshold, the risk of loan defaults increase and the burden of repayment falls on taxpayers and more successful borrowers [@BaumSandy2016SLPa].

This report will attempt to analyse the risks of HELP on its long-term sustainability through answering the questions of:

1.	<i>Alana’s</i>

2.	<i>Alana’s</i>

3.	<i>What was the distribution of taxable (HELP repayment) income of HELP debtors in Australian States and Territories in the 2021-22 financial year?</i>

4.	<i>What trend can be observed of the median and mean taxable (HELP repayment) income of HELP debtors in Australia over time?</i>

5.	<i>Chenxi’s</i>

6.	<i>Chenxi’s</i>


# Analysis

## Outstanding Debt and Debtors by Year, Age and Sex {#yearsection}

### How does total outstanding debt and number of debtors change from the financial years 2005-06 to 2021-22? {#yearsection}

```{r loading, include=FALSE, message=FALSE, warning=FALSE}

table_8 <- read.csv("Data/HELP_DATA_TIDIED.csv")

```

```{r table1, message=FALSE, warning=FALSE, echo=FALSE}
# Group By Year 
table_8_year <- table_8 %>%
  group_by(Year) %>%
  summarise(`Total Number of Debtors (millions)` = sum(Num_Debtors)/1000000, 
            `Total Outstanding Debt (billions $)` = sum(Outstanding_Debt)/1000000000) 

# Create Table
kable(table_8_year, 
      caption = "Total Debtors (in millions) and Total Oustanding Debt (in billions)",
             digits = 2)
```

```{r figure1, warning=FALSE, message=FALSE, echo=FALSE, fig.align='center', fig.cap="Total outstanding debt and number of debtors from the financial years 2005-06 to 2021-22"}
totalOut_plot <- ggplot(data = table_8_year, aes(x = Year, y = `Total Outstanding Debt (billions $)`)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Total Oustanding Debt ($) in billions") + 
  labs(title = "a) Outstanding Debt Has Increased From 2005", subtitle = "Total Outstanding Debt From 2005-2021") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)

totalDebtors_plot <- ggplot(data = table_8_year, aes(x = Year, y = `Total Number of Debtors (millions)`)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Total Debtors in millions") + 
  labs(title = "b) Number of Debtors Has Increased From 2005", subtitle = "Total Number of Debtors From 2005-2021") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)


grid.arrange(totalOut_plot, totalDebtors_plot, nrow = 2)
```

The total number of debtors increased from 1.18 million in 2005 to 3 million in 2021.  The outstanding debt total was $12.39 billion in 2005 which rose to $74.39 billion in 2021. A rapid increase can be seen from 2010 to 2015, where total outstanding debt rose by $25.32 billion. Interestingly, a small decrease in debtors can be observed from 2018 to 2019 which may be a factor for the plateau in outstanding debt from years 2018-2020. 

### Average Debtors and Oustanding Debt by Sex {#sexsection}

```{r table2, echo=FALSE, message=FALSE, warning=FALSE}
# Group By Year 
table_8_sex <- table_8 %>%
  group_by(Sex) %>%
  summarise(`Mean Number of Debtors` = mean(Num_Debtors), 
            `Mean Outstanding Debt (in billions)` = mean(Outstanding_Debt)/1000000000)

kable(table_8_sex, 
      caption = "Average Debtors and Outstanding Debt By Sex from 2005-2021",
             digits = 2)
```

```{r figure2, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Total outstanding debt and number of debtors by sex from the financial years 2005-06 to 2021-22"}
table_8_sex_year <- table_8 %>%
  group_by(Sex, Year) %>%
  summarise(mean_number_Debtors = mean(Num_Debtors), 
            mean_outstandingDebt_billions = mean(Outstanding_Debt)/1000000000)


plot3 <- ggplot(data = table_8_sex_year, aes(x = Year, y = mean_number_Debtors, col = Sex)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Mean Number of Debtors") + 
  labs(title = "a) There is an Increase in Both Male and Female Debtors", subtitle = "Females on average are more likely to be in debt") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)


plot4 <- ggplot(data = table_8_sex_year, aes(x = Year, y = mean_outstandingDebt_billions, col = Sex)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Mean Outstanding Debt ($) in billions") + 
  labs(title = "b) Females Have a Greater Outstanding Debt than Males", subtitle = "There is an increasing gap between outstanding debt by sex") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)

grid.arrange(plot3, plot4, nrow = 2)

```

Grouping mean debtors and outstanding debt by sex shows that there are significantly more female debtors than males. The reason is likely an increase in university enrolments, with female enrolments growing faster than male enrolments [@preston_gender_2023]. From 2005-2021, on average, 158, 046.2 females are in debt, compared to 103,717.2 males. This difference in average of 54,329 female debtors could factor for the $750 million difference in average outstanding debt. An increasing gap can be observed between the number of debtors and outstanding debt between females and males. 

### Average Debtors and Oustanding Debt by Age Group {#agesection}

```{r table3, echo=FALSE, message=FALSE, warning=FALSE}
# group by age 

table_8_age <- table_8 %>%
  group_by(Age_Range) %>%
  summarise(`Mean Number of Debtors (in thousands)` = mean(Num_Debtors)/1000, 
            `Mean Oustanding Debt (in millions)` = mean(Outstanding_Debt)/1000000)

kable(table_8_age, 
      caption = "Average Number of Debtors and Outstanding Debt by Age Group from 2005-2021",
             digits = 2)
```

```{r figure3, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Total outstanding debt and number of debtors by age from the financial years 2005-06 to 2021-22"}
# group by age and year 

table_8_age_year <- table_8 %>%
  group_by(Age_Range, Year) %>%
  filter(Age_Range != "Unknown") %>%
  summarise(mean_NumDebtors = mean(Num_Debtors), 
            mean_OustandingDebt = mean(Outstanding_Debt)) 

# kable(table_8_age_year)

plot5 <- ggplot(data = table_8_age_year, aes(x = Year, y = mean_NumDebtors, col = Age_Range)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Mean Number of Debtors") + 
  labs(title = "a) Ages 20-29 are the Largest Group of Debtors", subtitle = "Most age ranges have an increase in the number of debtors") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)

plot6 <- ggplot(data = table_8_age_year, aes(x = Year, y = mean_OustandingDebt, col = Age_Range)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Mean Outstanding Debt ($)") + 
  labs(title = "b) Ages 20-29 Have the Highest Outstanding Debt", subtitle = "There is a rapid increase for ages 30-39") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)

grid.arrange(plot5, plot6, nrow = 2)
```

The age group with the greatest number of debtors is 20-29. Excluding the group for unknown age, 70+ has the lowest number of debtors. The same pattern can be observed for mean outstanding debt. Younger age groups have a higher risk of debt than older age groups with the exception of 0-19. An increase in the number of debtors and outstanding debt can be seen for all age groups.  From 2018-2021, there is a sudden drop in number of debtors for 0–19-year-olds, which also results in a drop in outstanding debt. 

## Income and Outstanding Debt {#income}

<b> What was the distribution of taxable (HELP repayment) income of HELP debtors in Australian States and Territories in the 2021-22 financial year? </b>

```{r distribution, echo=FALSE, message=FALSE, warning=FALSE}

# Select columns and filter data
income_dist <- table_7 %>%
  select('Financial year', 'HELP repayment income', 'ACT', 'NSW', 'NT', 'QLD', 'SA', 'TAS', 'VIC', 'WA') %>%
  filter(`Financial year` == "2021–22") %>%
  filter(`HELP repayment income` != "Unknown")

# Pivot data
pivoted_dist <- income_dist %>%
  pivot_longer(
    cols = c('ACT', 'NSW', 'NT', 'QLD', 'SA', 'TAS', 'VIC', 'WA'),
    names_to = "State",
    values_to = "Count")

# Create factor to assign order to char type column
income_factor = c("Up to $9,999", 
                  "$10,000 to $19,999", 
                  "$20,000 to $29,999", 
                  "$30,000 to $39,999", 
                  "$40,000 to $49,999", 
                  "$50,000 to $59,999", 
                  "$60,000 to $69,999", 
                  "$70,000 to $79,999", 
                  "$80,000 to $89,999", 
                  "$90,000 to $99,999",
                  "$100,000 or more",
                  "$100,000 to $109,999", 
                  "$110,000 to $119,999", 
                  "$120,000 to $129,999", 
                  "$130,000 or more")


# Create plot
dist_plot <- ggplot(data = pivoted_dist, aes(x = factor(`HELP repayment income`, level = income_factor), y = Count, fill = State)) +
  geom_bar(stat = "identity") +
  theme(panel.grid.major = element_line(color = "grey",
                                        size = 0.5,
                                        linetype = 3),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_y_continuous(name = "Count of Debtors", labels = label_comma(), 
                     breaks = seq(0, 350000, by = 50000)) +
  xlab("Income Group") +
  labs(title = "The highest number of HELP debtors earn between $20,000 and $29,999", 
       subtitle = "There is a postive skew in the distribution of HELP debtor income", 
       caption = "Data source: Data.gov.au")

```

```{r disttab, echo=FALSE, message=FALSE, warning=FALSE}

# Display table
kable(income_dist, 
      col.names = c("Financial Year","HELP Repayment Income", "ACT", "NSW", "NT", "QLD", "SA", "TAS", "VIC", "WA"), 
      caption = "Table of count of outstanding HELP debtors in each income group by State or Territory in 2021-22")

```

```{r distfig, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Distribution of HELP debtor income in Australian States and Territories"}

# Display plot
dist_plot

```

Table \@ref(tab:disttab) presents the counts of students with outstanding HELP debt in each income group, incremented by $10,000, in all States and Territories in the 2021-22 fiscal year. The distribution of debtors is depicted in Figure \@ref(fig:distfig). Across all income groups, NSW, QLD, and VIC have the highest number of debtors. There are 336,992 outstanding HELP debtors earning between $20,000 and $29,999, the highest of all income groups. The number of debtors diminishes gradually as the income group increases. The income group with the lowest number of outstanding debtors is $120,000 to $129,999, with 19,642 debtors. It is to be noted that the “$130,000 or more” income group is an exception as it does not observe the $10,000 increment rule.

<b> What trend can be observed of the median and mean taxable (HELP repayment) income of HELP debtors in Australia over time? </b>

```{r median, echo=FALSE, message=FALSE, warning=FALSE}

# Select columns and filter data
median_income <- table_7 %>%
  select('Financial year', 'HELP repayment income', 'ACT', 'NSW', 'NT', 'QLD', 'SA', 'TAS', 'VIC', 'WA') %>%
  filter(`HELP repayment income` != "Unknown")

# Pivot data
median_pivot <- median_income %>%
  pivot_longer(
    cols = c('ACT', 'NSW', 'NT', 'QLD', 'SA', 'TAS', 'VIC', 'WA'),
    names_to = "State",
    values_to = "Count")

# Change income column from char type to numerical
median_pivot <- mutate(median_pivot, `HELP repayment income` = case_when(
    `HELP repayment income` == "Up to $9,999" ~ 9999,
    `HELP repayment income` == "$10,000 to $19,999" ~ 19999, 
    `HELP repayment income` == "$20,000 to $29,999" ~ 29999, 
    `HELP repayment income` == "$30,000 to $39,999" ~ 39999, 
    `HELP repayment income` == "$40,000 to $49,999" ~ 49999, 
    `HELP repayment income` == "$50,000 to $59,999" ~ 59999, 
    `HELP repayment income` == "$60,000 to $69,999" ~ 69999, 
    `HELP repayment income` == "$70,000 to $79,999" ~ 79999, 
    `HELP repayment income` == "$80,000 to $89,999" ~ 89999, 
    `HELP repayment income` == "$90,000 to $99,999" ~ 99999, 
    `HELP repayment income` == "$100,000 or more" ~ 100000,
    `HELP repayment income` == "$100,000 to $109,999" ~ 109999,
    `HELP repayment income` == "$110,000 to $119,999" ~ 119999, 
    `HELP repayment income` == "$120,000 to $129,999" ~ 129999, 
    `HELP repayment income` == "$130,000 or more" ~ 130000,
    is.na(`HELP repayment income`) ~ 0
    ))

# Create new variables
median_pivoted <- median_pivot %>%
  group_by(`Financial year`) %>%
  summarise(
    `Count of Debtors` = sum(Count),
    `Median Income` = sort(rep(`HELP repayment income`,Count))[length(rep(`HELP repayment income`,Count))/2],
    `Mean Income` = sum(sort(rep(`HELP repayment income`,Count)))/length(sort(rep(`HELP repayment income`,Count)))
  )

# Change median income column back to char type
median_pivoted <- mutate(median_pivoted, `Median Income` = case_when(
    `Median Income` == 9999 ~ "Up to $9,999",
    `Median Income` == 19999 ~ "$10,000 to $19,999", 
    `Median Income` == 29999 ~ "$20,000 to $29,999", 
    `Median Income` == 39999 ~ "$30,000 to $39,999", 
    `Median Income` == 49999 ~ "$40,000 to $49,999", 
    `Median Income` == 59999 ~ "$50,000 to $59,999", 
    `Median Income` == 69999 ~ "$60,000 to $69,999", 
    `Median Income` == 79999 ~ "$70,000 to $79,999", 
    `Median Income` == 89999 ~ "$80,000 to $89,999", 
    `Median Income` == 99999 ~ "$90,000 to $99,999",
    `Median Income` == 100000 ~ "$100,000 or more",
    `Median Income` == 109999 ~ "$100,000 to $109,999", 
    `Median Income` == 119999 ~ "$110,000 to $119,999", 
    `Median Income` == 129999 ~ "$120,000 to $129,999", 
    `Median Income` == 130000 ~ "$130,000 or more"
    ))

# Change mean income column back to char type
median_pivoted2 <- mutate(median_pivoted, `Mean Income` = case_when(
    `Mean Income` <= 9999 ~ "Up to $9,999",
    `Mean Income` <= 19999 ~ "$10,000 to $19,999", 
    `Mean Income` <= 29999 ~ "$20,000 to $29,999", 
    `Mean Income` <= 39999 ~ "$30,000 to $39,999", 
    `Mean Income` <= 49999 ~ "$40,000 to $49,999", 
    `Mean Income` <= 59999 ~ "$50,000 to $59,999", 
    `Mean Income` <= 69999 ~ "$60,000 to $69,999", 
    `Mean Income` <= 79999 ~ "$70,000 to $79,999", 
    `Mean Income` <= 89999 ~ "$80,000 to $89,999", 
    `Mean Income` <= 99999 ~ "$90,000 to $99,999",
    `Mean Income` == 100000 ~ "$100,000 or more",
    `Mean Income` <= 109999 ~ "$100,000 to $109,999", 
    `Mean Income` <= 119999 ~ "$110,000 to $119,999", 
    `Mean Income` <= 129999 ~ "$120,000 to $129,999", 
    `Mean Income` <= 130000 ~ "$130,000 or more"
    ))

# Create plot
median_plot <- ggplot(median_pivoted2, aes(x = `Financial year`)) +
  geom_line(aes(y = `Median Income`, group = 1, color = "Median")) +
  geom_point(aes(y = `Median Income`, group = 1, color = "Median")) +
  geom_line(aes(y = `Mean Income`, group = 1, color = "Mean")) +
  geom_point(aes(y = `Mean Income`, group = 1, color = "Mean")) +
  scale_color_manual(values = c(Median = "steelblue", Mean = "darkred"), 
                     labels = c(Median = "Median income", Mean = "Mean income")) +
  labs(title = "The median and mean income group of HELP debtors in 2021 was \n$40,000 to $49,999 and $50,000 to $59,999 respectively", 
       subtitle = "There has been a positive increase in median and mean taxable income over time", 
       caption = "Data source: Data.gov.au") +
  ylab("Income Group") +
  xlab("Year") +
  guides(col = guide_legend("Legend")) +
  theme(panel.grid.major = element_line(color = "grey",
                                        size = 0.5,
                                        linetype = 3),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

```{r mediantab, echo=FALSE, message=FALSE, warning=FALSE}

# Display the table
kable(median_pivoted2, 
      col.names = c("Financial Year", "Count of Outstanding Debtors", "Median Income", "Mean Income"), 
      caption = "Table of the median and mean taxable income group of outstanding HELP debtors in Australia from 2009-10 to 2021-22")

```

```{r medianfig, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Median and mean income of HELP debtors in Australia from 2009 to 2021"}

# Display the plot
median_plot

```

Table \@ref(tab:mediantab) presents the median and mean (average) income of HELP debtors in Australia from 2009 to 2021 and the trends are illustrated in Figure \@ref(fig:medianfig). The median and average income follow the same positive trend, increasing overall by one income group ($10,000) since 2009. The median typically sits below the average by one income group. In 2009, the median income group was $30,000 to $39,999 and the mean income group was $40,000 to $49,999. By 2021, the median had increased to $40,000 to $49,999 and the average to $50,000 to $59,999. The lowest median and mean income group of HELP debtors were observed in the 2010-11 to 2012-13 and 2015-16 fiscal years, at $20,000 to $29,999 and $30,000 to $39,999 respectively.

## Recoverability of HELP Debt Over Time

In order to investigate whether the Australian Government is able to recover debt issued under the HELP program, cumulative data available from FY07 to FY22 have been translated to annual increase for new debt incurred (including indexation) and repayments received (including compulsory repayments, voluntary repayments and bonus on voluntary repayments which ceased in FY17).

<b> Annual Debt Incurred vs Repayments </b>

``` {r DebtRepaymentTable, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap="TBC"}
table_1 <- add_column(table_1,'Financial Year'=as.numeric(substr(table_1$Year,1,4))+1)
table_1 <- arrange(table_1,desc(table_1[9]))
table_1 <- add_column(table_1,'Annual Incurred (incl. Indexation)'=(table_1$`Total incurred debt ($)`+table_1$`Indexation ($)`)-lag(table_1$`Total incurred debt ($)`+table_1$`Indexation ($)`,order_by=table_1[2]))
table_1 <- add_column(table_1,'Annual Repayment' = (table_1$`Compulsory repayments ($)`+ifelse(table_1$`OS Levy Compulsory repayments ($)`=="n/a",0,as.numeric(table_1$`OS Levy Compulsory repayments ($)`))+table_1$`Voluntary repayments ($)`+table_1$`Bonus on voluntary repayments ($)`)-lag(table_1$`Compulsory repayments ($)`++ifelse(table_1$`OS Levy Compulsory repayments ($)`=="n/a",0,as.numeric(table_1$`OS Levy Compulsory repayments ($)`))+table_1$`Voluntary repayments ($)`+table_1$`Bonus on voluntary repayments ($)`,order_by=table_1[2]))
table_1 <- add_column(table_1, 'Annual Repayment Ratio'=table_1$`Annual Repayment`/table_1$`Annual Incurred (incl. Indexation)`)
table_1 <- arrange(table_1,table_1$`Financial Year`)
tableDebtRepayment <- table_1[-c(1),c(10:13)]
kable(tableDebtRepayment,caption = "Annual New Debt Incurred and Repayments")

```
``` {r DebtRepaymentGraph, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Annual New Debt Incurred and Repayments"}
graphDebtRepayment = ggplot(table_1)+
  geom_line(aes(x=table_1$`Financial Year`, y=table_1$`Annual Incurred (incl. Indexation)`, color="New Debt Incurred"))+
  geom_line(aes(x=table_1$`Financial Year`, y=table_1$`Annual Repayment`, color="Repayment"))+
  geom_line(aes(x=table_1$`Financial Year`, y=table_1$`Annual Repayment Ratio`*8000000000, color="Repayment Ratio (RHS)"), size=3)+
  scale_x_continuous(breaks=seq(2007,2022,1))+
  scale_y_continuous("Annual New Debt and Payment ($)",sec.axis=sec_axis(~./8000000000,name="Annual Payment to New Debt Ratio (%)",labels = percent), labels = comma)+
  xlab("Financial Year")+
  labs(title="Annual Debt Incurred vs Repayment",subtitle="Increase in repayment ratio from 2016 indicating improved debt recoverability", color="Legend", caption="Data source: Data.gov.au")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
graphDebtRepayment
```

Graph \@ref(fig:DebtRepaymentGraph) shows significant increase in new debts under the program each year, start at below \$3 billion in FY07 and increasing to over \$10 billion per year in FY18 and FY22. Annual increase in repayments have occurred at a slower rate, but follows an exponential trajectory to FY22. Repayment to new debt ratio have been calculated as primary indicator for recoverability of debt issued under the scheme. Between FY07 to FY16, this indicators shows a concerning trend as it decreased from 38% to 21% as new debt outpaced repayments. However this have recovered substantially in recent years (reaching >100% in FY20 which is an outlier due to low new debt in the year) which means that even as greater amounts are lent under this program year on year, it  still appears to be a sustainable one as annual repayments received have grown steadily over the past 6 years.

<b> Debt Write Off Ratio </b>

``` {r DebtRatioTable, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap="TBC"}
table_1 <- add_column(table_1,'Annual Debt Writeoff' =table_1$`Value of debts writen off ($)`-lag(table_1$`Value of debts writen off ($)`,order_by=table_1[2]))
table_1 <- add_column(table_1,'Debt Writeoff Ratio' =table_1$`Annual Debt Writeoff`/table_1$`Annual Incurred (incl. Indexation)`)
tableDebtWriteoff <- table_1[-c(1),c(10,15)]
kable(tableDebtWriteoff,caption = "Debt Writeoff as % of Annual New Debt")
```

``` {r DebtWriteoffGraph, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Debt Writeoff as % of Annual New Debt"}
graphDebtWriteoff = ggplot(table_1)+
  geom_line(aes(x=table_1$`Financial Year`, y=table_1$`Debt Writeoff Ratio`, color="Debt Writeoff %"))+
  geom_point(aes(x=table_1$`Financial Year`, y=table_1$`Debt Writeoff Ratio`, color="Debt Writeoff %"))+
  scale_x_continuous(breaks=seq(2007,2022,1))+
  scale_y_continuous("Annual New Debt and Payment ($)", labels = percent)+
  xlab("Financial Year")+
  labs(title="Debt Writeoff %",subtitle="Higher debt write % in recent years compared to the early 2000s",color="Legend", caption="Data source: Data.gov.au")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
graphDebtWriteoff
```
In table \@ref(fig:DebtWriteoffGraph) Annual debt writeoff as percentage of new debts is similar to the measure of loss rate measured for major banks. For the year 2019, average loss rate across the 3 major banks (ANZ, CBA, NAB) averages 0.15%. The calculated HELP debt writeoff % for FY is 0.46% [@pwc2019]. While substantially higher than the private banks, the HELP debt writeoff % is not unreasonable given the difference in nature compared to a commercial loan.

# Conclusion

Based on the analysis in \@ref(income), most debtors earn below the average income, while the number of debtors earning above the average diminishes as the income group increases. The skewed distribution in Figure \@ref(fig:distfig) represents the government’s risk in recovering HELP debt in 2021. Up to $20,000 to $29,999, the risk increases as many debtors do not meet the minimum repayment threshold. However, beyond $30,000, as income increases, the government's risk decreases as more repayments are made through the tax system. The increase of median and mean income observed in Figure \@ref(fig:medianfig) indicate an overall inflation of debtor income since 2009, meaning an improved ability to repay the HELP debt. This observation is supported by Figure \@ref(fig:DebtRepaymentGraph) which shows significant increase in annual payments received from 2016 reflective of increased debtor income and being pushed into higher repayment brackets.

The number of debtors and outstanding debt has steadily increased since 2005. From 2005 to 2021, there has been an increase of 1.82 million people in debt. During this same period, there has been an increase in outstanding debt of $62 billion \@ref(yearsection). Females are more likely to have HELP debt and there is an increasing gap between males and females in debt \@ref(sexsection). It appears with age, the risk of student debt decreases. All age groups increased in debtors and outstanding debt for 2005-2021, except for the age group 0-19, which saw a sudden drop in 2018-2021 \@ref(agesection). Overall, Student debt levels are rising, along with a rise in older people with outstanding debt, especially women [@preston_gender_2023]. 

However, this analysis is limited by the data not accounting for the size of HELP debt incurred and how progressed the debtor is in their career. Further progressed careers and larger debts which may indicate higher education further both may increase the level of income earned, while newer graduates may be in a lower income group.

# References

<div id="refs"></div>

# Appendix

