---
title: Analysis of Australia's Higher Education Loan Program
  (HELP) Statistics
author: "Alana Dabelstein, Anh-Thu Hoang, Chenxi Zhang"
date: "01/06/2023"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: paper
    fig_width: 7
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Loading Libraries
library(readxl)

# Loading Data
table_1 <- read_xlsx("Data/help-statistics-2021-2022.xlsx", sheet = 2, skip = 2, col_names = TRUE)
table_2 <- read_xlsx("Data/help-statistics-2021-2022.xlsx", sheet = 3, skip = 2, col_names = TRUE)
table_3 <- read_xlsx("Data/help-statistics-2021-2022.xlsx", sheet = 4, skip = 2, col_names = TRUE)
table_4 <- read_xlsx("Data/help-statistics-2021-2022.xlsx", sheet = 5, skip = 2, col_names = TRUE)
table_7 <- read_xlsx("Data/help-statistics-2021-2022.xlsx", sheet = 8, skip = 2, col_names = TRUE)
```

## Introduction

Educational attainment is associated with higher labour force participation, higher income, greater wealth, lower financial stress and healthier behaviours (OECD, 2018). Despite this, individuals, many individuals have debt-averse attitudes which deter them from applying to university (Callender & Mason, 2017). Fees for tertiary education have grown fast over time whereas there has been a flat growth in the wages of young adults aged 25-34 (Preston, 2023). Enrolments and the number of people with outstanding HELP debts has steadily increased (Universities Australia, 2022). 

## Analysis

### Alana's Part

## Outstanding Debt and Debtors by Year, Age and Sex {#yearsection}

### How does total outstanding debt and number of debtors change from the financial years 2005-06 to 2021-22? {#yearsection}

```{r loading, include=FALSE, message=FALSE, warning=FALSE}
table_8 <- read.csv("Data/HELP_DATA_TIDIED.csv")

library(tidyverse)
library(gridExtra)
library(knitr)
library(bookdown)
library(eeptools)
```

```{r table1, message=FALSE, warning=FALSE, include=FALSE}
# Group By Year 
table_8_year <- table_8 %>%
  group_by(Year) %>%
  summarise(`Total Number of Debtors (millions)` = sum(Num_Debtors)/1000000, 
            `Total Outstanding Debt (billions $)` = sum(Outstanding_Debt)/1000000000) 

# Create Table
kable(table_8_year, 
      caption = "Total Debtors (in millions) and Total Oustanding Debt (in billions)",
             digits = 2)
```

```{r figure1, warning=FALSE, message=FALSE, include=FALSE, fig.align='center', fig.cap="Total outstanding debt and number of debtors from the financial years 2005-06 to 2021-22"}
totalOut_plot <- ggplot(data = table_8_year, aes(x = Year, y = `Total Outstanding Debt (billions $)`)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Total Oustanding Debt ($) in billions") + 
  labs(title = "a) Outstanding Debt Has Increased From 2005", subtitle = "Total Outstanding Debt From 2005-2021") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)

totalDebtors_plot <- ggplot(data = table_8_year, aes(x = Year, y = `Total Number of Debtors (millions)`)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Total Debtors in millions") + 
  labs(title = "b) Number of Debtors Has Increased From 2005", subtitle = "Total Number of Debtors From 2005-2021") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)


grid.arrange(totalOut_plot, totalDebtors_plot, nrow = 2)
```

The total number of debtors increased from 1.18 million in 2005 to 3 million in 2021.  The outstanding debt total was $12.39 billion in 2005 which rose to $74.39 billion in 2021. A rapid increase can be seen from 2010 to 2015, where total outstanding debt rose by $25.32 billion. Interestingly, a small decrease in debtors can be observed from 2018 to 2019 which may be a factor for the plateau in outstanding debt from years 2018-2020. 

### Average Debtors and Oustanding Debt by Sex {#sexsection}

```{r table2, include=FALSE, message=FALSE, warning=FALSE}
# Group By Year 
table_8_sex <- table_8 %>%
  group_by(Sex) %>%
  summarise(`Mean Number of Debtors` = mean(Num_Debtors), 
            `Mean Outstanding Debt (in billions)` = mean(Outstanding_Debt)/1000000000)

kable(table_8_sex, 
      caption = "Average Debtors and Outstanding Debt By Sex from 2005-2021",
             digits = 2)
```

```{r figure2, include=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
table_8_sex_year <- table_8 %>%
  group_by(Sex, Year) %>%
  summarise(mean_number_Debtors = mean(Num_Debtors), 
            mean_outstandingDebt_billions = mean(Outstanding_Debt)/1000000000)


plot3 <- ggplot(data = table_8_sex_year, aes(x = Year, y = mean_number_Debtors, col = Sex)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Mean Number of Debtors") + 
  labs(title = "a) There is an Increase in Both Male and Female Debtors", subtitle = "Females on average are more likely to be in debt") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)


plot4 <- ggplot(data = table_8_sex_year, aes(x = Year, y = mean_outstandingDebt_billions, col = Sex)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Mean Outstanding Debt ($) in billions") + 
  labs(title = "b) Females Have a Greater Outstanding Debt than Males", subtitle = "There is an increasing gap between outstanding debt by sex") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)

grid.arrange(plot3, plot4, nrow = 2)

```

Grouping mean debtors and outstanding debt by sex shows that there are significantly more female debtors than males. The reason is likely an increase in university enrolments, with female enrolments growing faster than male enrolments (Preston, 2023). From 2005-2021, on average, 158, 046.2 females are in debt, compared to 103,717.2 males. This difference in average of 54,329 female debtors could factor for the $750 million difference in average outstanding debt. An increasing gap can be observed between the number of debtors and outstanding debt between females and males. 

### Average Debtors and Oustanding Debt by Age Group {#agesection}

```{r table3, include=FALSE, message=FALSE, warning=FALSE}
# group by age 

table_8_age <- table_8 %>%
  group_by(Age_Range) %>%
  summarise(`Mean Number of Debtors (in thousands)` = mean(Num_Debtors)/1000, 
            `Mean Oustanding Debt (in millions)` = mean(Outstanding_Debt)/1000000)

kable(table_8_age, 
      caption = "Average Number of Debtors and Outstanding Debt by Age Group from 2005-2021",
             digits = 2)
```

```{r figure3, include=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
# group by age and year 

table_8_age_year <- table_8 %>%
  group_by(Age_Range, Year) %>%
  filter(Age_Range != "Unknown") %>%
  summarise(mean_NumDebtors = mean(Num_Debtors), 
            mean_OustandingDebt = mean(Outstanding_Debt)) 

# kable(table_8_age_year)

plot5 <- ggplot(data = table_8_age_year, aes(x = Year, y = mean_NumDebtors, col = Age_Range)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Mean Number of Debtors") + 
  labs(title = "a) Ages 20-29 are the Largest Group of Debtors", subtitle = "Most age ranges have an increase in the number of debtors") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)

plot6 <- ggplot(data = table_8_age_year, aes(x = Year, y = mean_OustandingDebt, col = Age_Range)) + 
  geom_point(alpha = 0.8) + 
  xlab("Year") + 
  ylab("Mean Outstanding Debt ($)") + 
  labs(title = "b) Ages 20-29 Have the Highest Outstanding Debt", subtitle = "There is a rapid increase for ages 30-39") +
  theme_minimal() +  
  geom_smooth()  + 
  scale_y_continuous(labels = scales::comma)

grid.arrange(plot5, plot6, nrow = 2)
```

The age group with the greatest number of debtors is 20-29. Excluding the group for unknown age, 70+ has the lowest number of debtors. The same pattern can be observed for mean outstanding debt. Younger age groups have a higher risk of debt than older age groups with the exception of 0-19. An increase in the number of debtors and outstanding debt can be seen for all age groups.  From 2018-2021, there is a sudden drop in number of debtors for 0–19-year-olds, which also results in a drop in outstanding debt. 


### Anh-Thu's Part

### Chenxi's Part

## Conclusion

The number of debtors and outstanding debt has steadily increased since 2005. From 2005 to 2021, there has been an increase of 1.82 million people in debt. During this same period, there has been an increase in outstanding debt of $62 billion \@ref(yearsection). Females are more likely to have HELP debt and there is an increasing gap between males and females in debt \@ref(sexsection). It appears with age, the risk of student debt decreases. All age groups increased in debtors and outstanding debt for 2005-2021, except for the age group 0-19, which saw a sudden drop in 2018-2021 \@ref(agesection). Overall, Student debt levels are rising, along with a rise in older people with outstanding debt, especially women  (Preston, 2023).  

## References

Callender, C., & Mason, G. (2017). Does Student Loan Debt Deter Higher Education Participation? New Evidence from England. Sage Journals, 671(1).

OECD. (2018). Benefits of educational attainment. Department of Education. Australian Government.

Preston, A. (2023). Gender and Australia’s Higher Education Loan Program. The University of Western Australia, Department of Economics UWA Business School. Women in Social & Economic Research (WiSER).

Universities Australia. (2022). 2022 Higher Education Facts and Figures. Universities Australia.


## Appendix
